The non-random assembly of indirect interactions in  \nplant-pollinator networks

### 2022-10-07 Manuscript/Main_text/Main_text_Journal_of_Animal_Ecol.Rmd
supp mat update

__Jose B. Lanuza^1,2^__  | __Alfonso Allen-Perkins^2,3^__ | __Ignasi Bartomeus^2^__

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organised is not trivial since they go beyond pairwise interactions, but may get blurred when considering global network descriptors. Recent work has shown that the meso-scale, the intermediate level of network complexity between the species and the global network, can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species and when these network subgraphs differ statistically from a benchmark, they are often referred to as "network motifs". Although motifs can capture relevant ecological information of species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some network subgraphs are consistently under- or over-represented, suggesting the presence of worldwide network motifs in plant-pollinator networks. In addition, we found a higher proportion of densely connected network subgraphs that, based on previous findings, could reflect that species relative abundances are the main driver shaping the structure of the meso-scale on plant-pollinator communities. Moreover, we found that distinct subgraph positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different groups of animal and plant species representing their main life history strategies (i.e., functional groups). For instance, we found that the functional group of ‘bees’ was over-represented in subgraph positions with a lower number of indirect interactions in contrast to the rest of floral visitors groups. Finally, we show that the observed functional group combinations within a subgraph can not be retrieved from their expected probabilities (i.e., joint probability distributions), indicating that plant and floral visitor associations within subgraphs are not random either. Therefore, our results highlight the presence of common motifs in plant-pollinator communities that are formed by a non-random association of plants and floral visitors functional groups.

@@

Indirect interactions, the mediated effect between two species by a third species, are ubiquitous in natural systems [@strauss1991; @holt1994; @wootton1994] and play a key role in shaping community dynamics [@werner2003; @walsh2013]. Although some of the most studied indirect interactions are the ones caused by trophic cascades with consumer-resource interactions [@abrams1996; @schmitz2001], previous studies often focus on antagonistic interactions and overlook mutualistic ones [@knight2005]. On the contrary, plant-pollinator research has traditionally investigated direct interactions but neglected indirect effects such as competitive or facilitative processes between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Conventional analytical tools in bipartite networks often discard relevant information about indirect interactions by condensing the information into single indices either at species or network level [@cirtwill2018; @simmons2019]. However, as previously explored with three species modules in unipartite food webs [@allesina2008; @stouffer2010], the emergent framework of bipartite subgraphs has proven to be a powerful analytical tool to better understand plant-pollinator community dynamics by capturing both direct and indirect effects [@simmons2019]. These network subgraphs that constitute the meso-scale are often referred to as the building blocks of the network and when their numbers occur at higher frequencies than in randomised networks, those are generally defined as ‘network motifs’ [@milo2002]. Indeed, the analysis of motifs in simulated plant-pollinator networks has revealed that the different drivers that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect effects [@simmons2020]. However, the global patterns of indirect interactions in real plant-pollinator networks and their ecological consequences remain unexplored (e.g., over- and under- represented network subgraphs).

@@

Here we evaluate the interaction patterns of the meso-scale and its ecological implications in plant-pollinator networks with the help of 60 empirical networks from 18 different studies distributed across 14 different countries. For this, plants and pollinators were divided into functional groups that summarise their main life history strategies and the different networks were split into all their subgraph elements up to five nodes. First, based on previous knowledge showing the non random assembly of macro-scale mutualistic networks [@bascompte2003] and food webs [@dunne2002], we test the hypothesis that  the building blocks that constitute the meso-scale in plant-pollinator networks have a non-random structure of interactions with over- and under-represented subgraphs (i.e., motifs) that represent contrasting ecological roles. Second, because species relative abundances (i.e., neutral effects) are one of the main drivers of network macro-struture [@krishna2008; @suweis2013; @bartomeus2016], we hypothesise that neutral effects may also govern the interaction patterns of the meso-scale in plant-pollinator networks. Finally, we explore if plants and pollinators functional groups are associated with subgraph positions that imply distinct ecological roles and if there are over- and under-represented functional group combinations within the different network subgraphs.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the different plant and floral visitor species from these networks into functional groups that summarise their main life-history strategies. We calculated the normalised subgraph frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we investigated the ecological implications of their overall representation on the different subgraph positions, by comparing, on the one hand, the observed (normalised) frequencies of each group in the different subgraph positions with the frequencies of their simulated counterparts, and, on the other hand, the differences among groups per subgraph position in terms of their number of indirect interactions and specialisation. Finally, to understand if the different functional group combinations within a subgraph are structured, we compare if the observed functional group combinations from empirical networks can be retrieved from the joint probability distribution of the expected frequencies of the different plant and floral visitor groups within the network subgraph.

@@

We have compiled 60 plant-pollinator networks from 18 different studies (Table S1). All studies sampled plant-pollinator interactions in natural systems in a single flowering season and were selected based on wide geographical coverage (Fig. S1) and presence of interaction frequency as a measure of interaction strength. In total, there were 503 plant species, 1,111 floral visitors species and 6,248 pairwise interactions registered. For ease of data manipulation, plant and floral visitor species names were standardised with the help of the R package _taxize_ version _0.9.99_ [@chamberlain2020].

**Plant and floral visitor groups**

@@

We opted to calculate the plant functional groups within a larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the _FD_ package version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the _maptree_ package version _1.4-7_ [@white2009].

Second, floral visitors were aggregated into taxonomic groups (hereafter referred to as functional groups) based on taxonomic rank as done similarly in other plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomic complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into functional groups that represented approximately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other functional groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last functional group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

We calculated the frequency of all subgraphs up to five nodes (17 different subgraphs in total; see Fig. 1) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on subgraphs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network structures. Throughout the entire manuscript, floral visitors occupy the top nodes of the different network subgraphs and plants the bottom ones. To control for variation in network size, subgraph frequencies were normalised as a proportion of the total number of subgraphs within each subgraph class (i.e., 2-species, 3-species, 4-species and 5-species network subgraphs). In addition, we excluded 2-species (or nodes) subgraphs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different subgraphs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$), where $\overline{x}$ represents the mean of the average path length across subgraphs within the same group. In general terms, plant and floral visitor groups on more densely connected network subgraphs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected structures (e.g., medium-weak or weak).

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variability and provided highly similar or identical networks to the original ones. After extracting the subgraph frequencies from the simulated networks, for each subgraph type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed subgraph frequencies. Subgraphs whose percentile is close to 0 or 100 are under- or over-represented in empirical networks, respectively, and indicate the presence of network motifs as they differ statistically from the network structures generated with simulated networks. To summarise general patterns across networks, we used an intercept-only linear mixed model (LMM) per subgraph with the help of the _lme4_ package version _1.1-21_ [@bates2015], where the response variable was the observed subgraph percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average subgraph frequency, but controlling for the variation at the study level.

**Over- and under-represented groups on subgraph positions**

We calculated which plant and floral visitor functional groups were over- or under-represented in the different subgraph positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same subgraph size class (i.e., 2, 3, 4 or 5 species motifs). Then, we calculated the percentile of the observed position frequencies for each functional group and network, just like we did with subgraph frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per subgraph position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average subgraph frequency per species group, after controlling for the variation at the study level. Finally, we visualised over- and under-representation of plant and floral visitor groups on the different subgraph positions with the help of the _ComplexHeatmap_ package version _2.6.2_ [@gu2016].

To understand if the different plant and floral visitor functional groups are associated with subgraph positions that represent different network roles, we characterised the roles of each subgraph position for the different functional groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each subgraph position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given subgraph. Values of the vector $si$, range between 1, complete specialisation (1 partner within the subgraph) and 0 complete generalism (all possible partners within the subgraph). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($\theta_i$) for each motif position as $\theta_i=m_i-k_i-1$, where $m_i$ is the total number of nodes in a given motif, and $k_i$ represents the degree of the corresponding motif position, that is, the total number of links connected to such position. Thus, we considered all possible indirect interactions for each position. Finally, we analysed in a single model how specialisation and the number of indirect interactions were associated with the observed percentiles for each trophic guild and subgraph position (see paragraph above). For this, we fitted again a LMM model with the help of the _lme4_ package where we considered specialisation and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor subgraph position nested within study system.

**Over- and under-represented group combinations of subgraphs**

@@

All analyses and data manipulation were conducted in R _version 4.0.5_ [@R]. In addition, all model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualised with the _visreg_ package [@breheny2020] version 2.7.0.

@@

Most network subgraphs were under- or over-represented (close to the 1st and 99th percentile, respectively) in the comparison between empirical and simulated networks (Fig. 2). Network subgraphs 3, 5, 9, 10 and 14 were under-represented in empirical networks, that is, all were close to the 1st percentile and under the 25th percentile. Interestingly, four out of five of these motifs belonged to the largest path length classification (i.e., core-peripheral). In addition, network subgraphs 2, 6, 7, 16 and 17 were over-represented, all over the 75th percentile. In contrast to the under-represented motifs, over-represented motifs belonged to the two shortest path length groups (i.e., complete and fan). The remaining network subgraphs (i.e., 4, 8, 11, 12, 13, and 15) were between the 25th and 75th percentile.

@@

labels <- c("2" = "*Motif 2* <img src='../../Images/motif_2.png' width='22'/>" ,
          "3"= "*Motif 3* <img src='../../Images/motif_3.png' width='22'/>",
          "4"= "*Motif 4* <img src='../../Images/motif_4.png' width='22'/>",
          "5"= "*Motif 5* <img src='../../Images/motif_5.png' width='22'/>",
          "6"= "*Motif 6* <img src='../../Images/motif_6.png' width='22'/>",
          "7"= "*Motif 7* <img src='../../Images/motif_7.png' width='22'/>",
          "8"= "*Motif 8* <img src='../../Images/motif_8.png' width='22'/>",
          "9"= "*Motif 9* <img src='../../Images/motif_9.png' width='22'/>",
          "10"= "*Motif 10* <img src='../../Images/motif_10.png' width='22'/>",
          "11"= "*Motif 11* <img src='../../Images/motif_11.png' width='22'/>",
          "12"= "*Motif 12* <img src='../../Images/motif_12.png' width='22'/>",
          "13"= "*Motif 13* <img src='../../Images/motif_13.png' width='22'/>",
          "14"= "*Motif 14* <img src='../../Images/motif_14.png' width='22'/>",
          "15"= "*Motif 15* <img src='../../Images/motif_15.png' width='22'/>",
          "16"= "*Motif 16* <img src='../../Images/motif_16.png' width='22'/>",
          "17"= "*Motif 17* <img src='../../Images/motif_17.png' width='22'/>")

@@

**Over- and under-represented groups on subgraph positions**

@@

The comparison of the plant and floral visitor functional group frequencies per subgraph position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (<25th) groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented groups, respectively. Notably, the differences across groups were more marked for floral visitors than for plant groups (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors groups on the different subgraph positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented groups than others, thus from most over- to under-represented functional groups on the different subgraph positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant groups we found over- and under-representation on different subgraph positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on subgraph positions that involved specialisation (i.e., lower specificity values) while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised subgraph positions with low number of indirect interactions (Fig. S8 A). Although plants did not show differences with null models on their frequencies, all plant groups but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised subgraph positions (Fig. S8 B). Second, regarding the number of indirect interactions, bees were over-represented in subgraph positions with low number of indirect interactions (Fig. S9 A) and the rest of floral visitor functional groups showed no differences across positions in comparison with simulated networks. For plants, all functional groups showed a slight trend to be over-represented on subgraph positions with higher number of indirect interactions (Fig. S9 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant functional groups, we found that both bees and tall plants with small unisexual flowers were the only groups with high frequencies on subgraph positions associated with low number interactions in comparison with the other floral visitor and plant functional groups (Figs S10 and S11).

```{r, echo=FALSE, message=FALSE, cache=FALSE,  warning=FALSE, fig.height=8,fig.width=8,fig.align = 'center', out.width="80%",fig.cap="\\textbf{Figure 3.} Heatmap indicating under- and over-representation of floral visitor and plant groups in the different subgraph positions (numbers on the right of each heatmap). The different subgraph positions are dividied by the average path length clasification determined by Simmons et al. (2020) and they are ordered from the most densely connected network structures (strong or complete) to the less connected ones (weak or core-peripheral). The superior dendrogram indicates the differences across groups with the more separated groups showing larger differences."}

@@

All data and code are publicly available at Github (https://github.com/JoseBSL/FunctionalMotifs) and Zenodo (https://doi.org/10.5281/zenodo.6489213).

@@

<div id ="refs"></div>

>>>>

**Jose B. Lanuza^1,2^** \| **Alfonso Allen-Perkins^2,3^** \| **Ignasi Bartomeus^2^**

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organised is not trivial since they go beyond pairwise interactions, but may get blurred when considering global network descriptors. Recent work has shown that the meso-scale, the intermediate level of network complexity between the species and the global network, can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species and when these network subgraphs differ statistically from a benchmark, they are often referred to as "network motifs". Although motifs can capture relevant ecological information of species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some network subgraphs are consistently under- or over-represented, suggesting the presence of worldwide network motifs in plant-pollinator networks. In addition, we found a higher proportion of densely connected network subgraphs that, based on previous findings, could reflect that species relative abundances are the main driver shaping the structure of the meso-scale on plant-pollinator communities. Moreover, we found that distinct subgraph positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different groups of animal and plant species representing their main life-history strategies (i.e., functional groups). For instance, we found that the functional group of 'bees' was over-represented in subgraph positions with a lower number of indirect interactions in contrast to the rest of floral visitors groups. Finally, we show that the observed functional group combinations within a subgraph can not be retrieved from their expected probabilities (i.e., joint probability distributions), indicating that plant and floral visitor associations within subgraphs are not random either. Therefore, our results highlight the presence of common motifs in plant-pollinator communities that are formed by a non-random association of plants and floral visitors functional groups.

@@

Indirect interactions, the mediated effect between two species by a third species, are ubiquitous in natural systems [@strauss1991; @holt1994; @wootton1994] and play a key role in shaping community dynamics [@werner2003; @walsh2013]. Although some of the most studied indirect interactions are the ones caused by trophic cascades with consumer-resource interactions [@abrams1996; @schmitz2001], previous studies often focus on antagonistic interactions and overlook mutualistic ones [@knight2005]. On the contrary, plant-pollinator research has traditionally investigated direct interactions but neglected indirect effects such as competitive or facilitative processes between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Conventional analytical tools in bipartite networks often discard relevant information about indirect interactions by condensing the information into single indices either at species or network level [@cirtwill2018; @simmons2019]. However, as previously explored with three species modules in unipartite food webs [@allesina2008; @stouffer2010], the emergent framework of bipartite subgraphs has proven to be a powerful analytical tool to better understand plant-pollinator community dynamics by capturing both direct and indirect effects [@simmons2019]. These network subgraphs that constitute the meso-scale are often referred to as the building blocks of the network and when their numbers occur at higher frequencies than in randomised networks, those are generally defined as 'network motifs' [@milo2002]. Indeed, the analysis of motifs in simulated plant-pollinator networks has revealed that the different drivers that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect effects [@simmons2020]. However, the global patterns of indirect interactions in real plant-pollinator networks and their ecological consequences remain unexplored (e.g., over- and under- represented network subgraphs).

@@

Here we evaluate the interaction patterns of the meso-scale and its ecological implications in plant-pollinator networks with the help of 60 empirical networks from 18 different studies distributed across 14 different countries. For this, plants and pollinators were divided into functional groups that summarise their main life-history strategies and the different networks were split into all their subgraph elements up to five nodes. First, based on previous knowledge showing the non random assembly of macro-scale mutualistic networks [@bascompte2003] and food webs [@dunne2002], we test the hypothesis that the building blocks that constitute the meso-scale in plant-pollinator networks have a non-random structure of interactions with over- and under-represented subgraphs (i.e., motifs) that represent contrasting ecological roles. Second, because species relative abundances (i.e., neutral effects) are one of the main drivers of network macro-struture [@krishna2008; @suweis2013; @bartomeus2016], we hypothesise that neutral effects may also govern the interaction patterns of the meso-scale in plant-pollinator networks. Finally, we explore if plants and pollinators functional groups are associated with subgraph positions that imply distinct ecological roles and if there are over- and under-represented functional group combinations within the different network subgraphs.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the different plant and floral visitor species from these networks into functional groups that summarise their main life-history strategies. We calculated the normalised subgraph frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we investigated the ecological implications of their overall representation on the different subgraph positions, by comparing, on the one hand, the observed (normalised) frequencies of each group in the different subgraph positions with the frequencies of their simulated counterparts, and, on the other hand, the differences among groups per subgraph position in terms of their number of indirect interactions and specialisation. Finally, to understand if the different functional group combinations within a subgraph are structured, we compare if the observed functional group combinations from empirical networks can be retrieved from the joint probability distribution of the expected frequencies of the different plant and floral visitor groups within network subgraphs.

@@

We have compiled 60 plant-pollinator networks from 18 different studies (Table S1). All studies sampled plant-pollinator interactions in natural systems in a single flowering season and were selected based on wide geographical coverage (Fig. S1) and presence of interaction frequency as a measure of interaction strength. In total, there were 503 plant species, 1,111 floral visitors species and 6,248 pairwise interactions registered. For ease of data manipulation, plant and floral visitor species names were standardised with the help of the R package *taxize* version *0.9.99* [@chamberlain2020].

@@

We opted to calculate the plant functional groups within a larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function *gowdis* with method *ward.D2* from the *FD* package version *1.0-12* [@laliberte2014]. Finally, we conducted hierarchical clustering with the function *hclust* from the *R* stats package version *4.0.5* and calculated the optimal number of clusters with the function *kgs* from the *maptree* package version *1.4-7* [@white2009].

Second, floral visitors were aggregated into taxonomic groups (hereafter referred to as functional groups) based on taxonomic rank as done similarly in other plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomic complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into functional groups that represented approximately the main life-history strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other functional groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last functional group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

We calculated the frequency of all subgraphs up to five nodes (17 different subgraphs in total; see Fig. 1) for each empirical network, by using the *bmotif* package version *2.0.2* [@bmotifsimmons2019]. We focused exclusively on subgraphs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network structures. Throughout the entire manuscript, floral visitors occupy the top nodes of the different network subgraphs and plants the bottom ones. To control for variation in network size, subgraph frequencies were normalised as a proportion of the total number of subgraphs within each subgraph class (i.e., 2-species, 3-species, 4-species and 5-species network subgraphs). In addition, we excluded 2-species (or nodes) subgraphs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different subgraphs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$), where $\overline{x}$ represents the mean of the average path length across subgraphs within the same group. In general terms, plant and floral visitor groups on more densely connected network subgraphs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected structures (e.g., medium-weak or weak).

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the *nullmodel* function and the *vaznull* model in the *bipartite* package version *2.14* [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variability and provided highly similar or identical networks to the original ones. After extracting the subgraph frequencies from the simulated networks, for each subgraph type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed subgraph frequencies. Subgraphs whose percentile is close to 0 or 100 are under- or over-represented in empirical networks, respectively, and indicate the presence of network motifs as they differ statistically from the network structures generated with simulated networks. To summarise general patterns across networks, we used an intercept-only linear mixed model (LMM) per subgraph with the help of the *lme4* package version *1.1-21* [@bates2015], where the response variable was the observed subgraph percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average subgraph frequency, but controlling for the variation at the study level.

**Over- and under-represented functional groups on subgraph positions**

We calculated which plant and floral visitor functional groups were over- or under-represented in the different subgraph positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same subgraph size class (i.e., 2, 3, 4 or 5 species motifs). Then, we calculated the percentile of the observed position frequencies for each functional group and network, just like we did with subgraph frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per subgraph position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average subgraph frequency per species group, after controlling for the variation at the study level. Finally, we visualised over- and under-representation of plant and floral visitor groups on the different subgraph positions with the help of the *ComplexHeatmap* package version *2.6.2* [@gu2016].

To understand if the different plant and floral visitor functional groups are associated with subgraph positions that represent different network roles, we characterised the roles of each subgraph position for the different functional groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each subgraph position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given subgraph. Values of the vector $si$, range between 1, complete specialisation (1 partner within the subgraph) and 0 complete generalism (all possible partners within the subgraph). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($\theta_i$) for each motif position as $\theta_i=m_i-k_i-1$, where $m_i$ is the total number of nodes in a given motif, and $k_i$ represents the degree of the corresponding motif position, that is, the total number of links connected to such position. Thus, we considered all possible indirect interactions for each position. Finally, we analysed in a single model how specialisation and the number of indirect interactions were associated with the observed percentiles for each trophic guild and subgraph position (see paragraph above). For this, we fitted again a LMM model with the help of the *lme4* package where we considered specialisation and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor subgraph position nested within study system.

@@

All analyses and data manipulation were conducted in R *version 4.0.5* [@R]. In addition, all model diagnostics were conducted with the help of the *Dharma* package [@hartig2017] version *0.3.3.0* and model outputs were visualised with the *visreg* package [@breheny2020] version 2.7.0.

@@

Most network subgraphs were under- or over-represented (close to the 1st and 99th percentile, respectively) in the comparison between empirical and simulated networks (Fig. 2). Network subgraphs 3, 5, 9, 10 and 14 were under-represented in empirical networks, that is, all were close to the 1st percentile and under the 25th percentile. Interestingly, four out of five of these motifs belonged to the largest path length classification (i.e., core-peripheral). In addition, network subgraphs 2, 6, 7, 16 and 17 were over-represented, all over the 75th percentile. In contrast to the under-represented network subgraphs, over-represented subgraphs belonged to the two shortest path length groups (i.e., complete and fan). The remaining network subgraphs (i.e., 4, 8, 11, 12, 13, and 15) were between the 25th and 75th percentile.

@@

labels <- c("2" = "*Subgraph 2* <img src='../../Images/motif_2.png' width='22'/>" ,
          "3"= "*Subgraph 3* <img src='../../Images/motif_3.png' width='22'/>",
          "4"= "*Subgraph 4* <img src='../../Images/motif_4.png' width='22'/>",
          "5"= "*Subgraph 5* <img src='../../Images/motif_5.png' width='22'/>",
          "6"= "*Subgraph 6* <img src='../../Images/motif_6.png' width='22'/>",
          "7"= "*Subgraph 7* <img src='../../Images/motif_7.png' width='22'/>",
          "8"= "*Subgraph 8* <img src='../../Images/motif_8.png' width='22'/>",
          "9"= "*Subgraph 9* <img src='../../Images/motif_9.png' width='22'/>",
          "10"= "*Subgraph 10* <img src='../../Images/motif_10.png' width='22'/>",
          "11"= "*Subgraph 11* <img src='../../Images/motif_11.png' width='22'/>",
          "12"= "*Subgraph 12* <img src='../../Images/motif_12.png' width='22'/>",
          "13"= "*Subgraph 13* <img src='../../Images/motif_13.png' width='22'/>",
          "14"= "*Subgraph 14* <img src='../../Images/motif_14.png' width='22'/>",
          "15"= "*Subgraph 15* <img src='../../Images/motif_15.png' width='22'/>",
          "16"= "*Subgraph 16* <img src='../../Images/motif_16.png' width='22'/>",
          "17"= "*Subgraph 17* <img src='../../Images/motif_17.png' width='22'/>")

@@

**Over- and under-represented functional groups on subgraph positions**

@@

The comparison of the plant and floral visitor functional group frequencies per subgraph position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (\<25th) functional groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented functional groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented functional groups, respectively. Notably, the differences across functional groups were more marked for floral visitors than for plants (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors functional groups on the different subgraph positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented functional groups than others, thus from most over- to under-represented functional groups on the different subgraph positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant functional groups we found over- and under-representation on different subgraph positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on subgraph positions that involved specialisation (i.e., lower specificity values) while the rest of functional groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised subgraph positions with low number of indirect interactions (Fig. S8 A). Although plants did not show differences with null models on their frequencies, all plant functional groups but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised subgraph positions (Fig. S8 B). Second, regarding the number of indirect interactions, bees were over-represented in subgraph positions with low number of indirect interactions (Fig. S9 A) and the rest of floral visitor functional groups showed no differences across positions in comparison with simulated networks. For plants, all functional groups showed a slight tendency to be over-represented on subgraph positions with higher number of indirect interactions (Fig. S9 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant functional groups, we found that both bees and tall plants with small unisexual flowers were the only functional groups with high frequencies on subgraph positions associated with low number interactions in comparison with the other floral visitor and plant functional groups (Figs S10 and S11).

```{r, echo=FALSE, message=FALSE, cache=FALSE,  warning=FALSE, fig.height=8,fig.width=8,fig.align = 'center', out.width="80%",fig.cap="\\textbf{Figure 3.} Heatmap indicating under- and over-representation of floral visitor and plant functional groups in the different subgraph positions (numbers on the right of each heatmap). The different subgraph positions are dividied by the average path length clasification determined by Simmons et al. (2020) and they are ordered from the most densely connected network structures (strong or complete) to the less connected ones (weak or core-peripheral). The superior dendrogram indicates the differences across functional groups with the more separated groups showing larger differences."}

@@

All data and code are publicly available at Github (<https://github.com/JoseBSL/FunctionalMotifs>) and Zenodo (<https://doi.org/10.5281/zenodo.6489213>).

@@

::: {#refs}
:::
